---
title: "Analyzing archaeological palimpsests with palimpsestr"
author: "Enzo Cocca"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing archaeological palimpsests with palimpsestr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

## Overview

**palimpsestr** provides a probabilistic framework for decomposing archaeological
palimpsests into latent depositional phases. The method integrates four domains
of evidence — spatial proximity, stratigraphic depth, chronological overlap,
and cultural similarity — to produce soft (probabilistic) phase assignments
and three diagnostic statistics:

- **SEI** (Stratigraphic Entanglement Index): measures pairwise depositional coherence
- **ESE** (Excavation Stratigraphic Energy): measures local depositional disruption
- **PDI** (Palimpsest Dissolution Index): global measure of phase separability

## Quick start

```{r quickstart}
library(palimpsestr)

# Simulate a palimpsest with 3 phases and 30% mixing
x <- archaeo_sim(n = 200, k = 3, mixing = 0.30, seed = 42)
head(x)
```

## Fitting the model

```{r fit}
fit <- fit_sef(
  x,
  k        = 3,
  tafonomy = "taf_score",
  context  = "context",
  seed     = 42
)

print(fit)
```

```{r summary}
summary(fit)
```

## Visualizing results

### Phase assignment

```{r phasefield, fig.cap="Dominant phase assignment. Point size reflects assignment confidence."}
plot_phasefield(fit)
```

### Entropy map

High entropy indicates areas where phase assignment is uncertain —
typically at the boundaries between phases or in heavily mixed zones.

```{r entropy, fig.cap="Spatial entropy map."}
plot_entropy(fit)
```

### Excavation Stratigraphic Energy

Elevated ESE values highlight zones of depositional disruption.

```{r energy, fig.cap="Excavation Stratigraphic Energy field."}
plot_energy(fit)
```

### SEI profile

```{r sei, fig.cap="Ranked local SEI values."}
plot_sei_profile(fit)
```

## Phase probability table

```{r phase_table}
pt <- as_phase_table(fit)
head(pt, 10)
```

## Intrusion detection

Artefacts with high entropy, high energy, and low SEI are flagged
as potentially intrusive.

```{r intrusions}
di <- detect_intrusions(fit)
di_top <- di[order(di$intrusion_prob, decreasing = TRUE), ]
head(di_top, 10)
```

## Selecting the number of phases

```{r compare_k}
ck <- compare_k(x, k_values = 2:6, tafonomy = "taf_score",
                context = "context", seed = 42)
print(ck)
```

The optimal K minimises BIC while maximising PDI.

## Using included demo datasets

The package ships with three simulated datasets representing different
degrees of palimpsest formation:

```{r demo_data, eval=FALSE}
data(demo_easy)        # 3 phases, 5% mixing
data(demo_moderate)    # 3 phases, 30% mixing
data(demo_compressed)  # 4 phases, 50% mixing
```

## GIS export

Results can be exported as `sf` objects for visualization in QGIS
or further spatial analysis.

```{r gis, eval=FALSE}
library(sf)

# Points with phase assignments
sf_pts <- as_sf_phase(fit, crs = 32632L)

# High-SEI links as lines
sf_links <- as_sf_links(fit, quantile_threshold = 0.90, crs = 32632L)

# Write to GeoPackage
st_write(sf_pts, "palimpsest_phases.gpkg", layer = "phases")
st_write(sf_links, "palimpsest_phases.gpkg", layer = "sei_links")
```

## Integration with PyArchInit

Results exported as GeoPackage can be loaded directly in QGIS
alongside PyArchInit layers. The `phase_diagnostic_table()` function
returns a data.frame compatible with CSV import into the PyArchInit
database.

```{r pyarchinit, eval=FALSE}
pdt <- phase_diagnostic_table(fit)
write.csv(pdt, "palimpsest_diagnostics.csv", row.names = FALSE)
```
