---
title: "palimpsestr: User Manual"
subtitle: "Complete Reference Guide v0.19.0"
author: "Enzo Cocca"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5, dpi = 120,
                      warning = FALSE, message = FALSE)
library(palimpsestr)
```

# Introduction

**palimpsestr** is an R package for the probabilistic decomposition of archaeological palimpsests using the Stratigraphic Entanglement Field (SEF) framework. It integrates spatial proximity, vertical distribution, chronological overlap, and cultural similarity to estimate latent depositional phases.

## Installation

```{r eval=FALSE}
# From GitHub
devtools::install_github("enzococca/palimpsestr")

# From local source
install.packages("palimpsestr_0.19.0.tar.gz", repos = NULL, type = "source")
```

## Required and optional packages

| Package | Status | Purpose |
|---------|--------|---------|
| stats, utils, graphics, grDevices | Required | Core R |
| ggplot2 | Suggested | Publication-quality plots (`gg_*` functions) |
| plotly | Suggested | Interactive visualization (`as_plotly()`) |
| sf | Suggested | GIS export (`as_sf_phase()`, `as_sf_links()`) |
| DBI, RSQLite, RPostgres | Suggested | Database import (`read_db()`) |

# Data Preparation

## Required columns

Every dataset must contain:

| Column | Type | Description |
|--------|------|-------------|
| `x` | numeric | Easting coordinate (metres) |
| `y` | numeric | Northing coordinate (metres) |
| `z` | numeric | Depth or elevation (metres) |
| `date_min` | numeric | Start of chronological interval |
| `date_max` | numeric | End of chronological interval |
| `class` | character | Material class (ceramic, lithic, bone, metal, etc.) |

## Optional columns

| Column | Type | Description |
|--------|------|-------------|
| `id` | any | Unique identifier for each find |
| `context` | character | Stratigraphic unit label (e.g., "SU_1") |
| `taf_score` | numeric (0--1) | Taphonomic disturbance score |
| `true_phase` | integer | Known phase (for validation only) |

## Notes on Z coordinates

**Each artifact needs its own z value.** If you don't have point-level depth measurements, you can assign the mean depth of the stratigraphic unit (US) to all finds within it:

```{r eval=FALSE}
# Example: assign US mean depth to finds
data$z <- ave(data$z_us, data$context, FUN = mean)
```

## Calibrated dates: rcarbon and OxCal

Calibrated radiocarbon or OxCal-modelled dates can be converted to the
`date_min`/`date_max`/`date_mid` columns expected by `fit_sef()`:

```{r eval=FALSE}
# From rcarbon
cal <- rcarbon::calibrate(x = c(2500, 2400), errors = c(30, 30))
chronology_from_rcarbon(cal)

# From OxCal (an oxcAAR result, or a data.frame of start/end ranges)
ranges <- data.frame(id = c("a", "b"), start = c(-700, -50), end = c(-600, 100))
chronology_from_oxcal(ranges)
```

## Simulating test data

```{r}
# Well-separated phases (low mixing)
easy <- archaeo_sim(n = 150, k = 3, seed = 1, mixing = 0.05)

# Moderate palimpsest
moderate <- archaeo_sim(n = 200, k = 3, seed = 1, mixing = 0.30)

# Compressed palimpsest (hard to resolve)
compressed <- archaeo_sim(n = 250, k = 4, seed = 1, mixing = 0.50)
```

Parameters:

- `n`: number of artifacts
- `k`: number of phases
- `seed`: random seed for reproducibility
- `mixing`: proportion of inter-phase mixing (0 = none, 1 = complete)

## Built-in datasets

The package includes four built-in datasets:

| Dataset | Finds | Phases | Mixing | Description |
|---------|-------|--------|--------|-------------|
| `demo_easy` | 150 | 3 | 5% | Well-separated phases |
| `demo_moderate` | 200 | 3 | 30% | Moderate palimpsest |
| `demo_compressed` | 250 | 4 | 50% | Heavily disturbed |
| `villa_romana` | 615 | -- | Real data | Poggio Gramignano Roman villa (VRPG 3004) |

```{r}
data(villa_romana)
str(villa_romana)
table(villa_romana$class)
# Core Analysis

## Fitting the SEF model: `fit_sef()`

```{r}
fit <- fit_sef(
  data     = moderate,
  coords   = c("x", "y", "z"),       # coordinate columns
  chrono   = c("date_min", "date_max"), # chronological range columns
  class    = "class",                  # material class column
  tafonomy = "taf_score",             # optional: taphonomic score
  context  = "context",               # optional: stratigraphic unit
  harris   = NULL,                    # optional: Harris Matrix penalty
  k        = 3,                       # number of phases
  weights  = c(ws = 1, wz = 1, wt = 1, wc = 1), # domain weights
  seed     = 42,                      # random seed
  em_iter  = 100,                     # max EM iterations
  em_tol   = 1e-5,                    # convergence tolerance
  n_init   = 5                        # number of random initialisations
)
print(fit)
```

### Parameters explained

| Parameter | Default | Description |
|-----------|---------|-------------|
| `k` | 3 | Number of depositional phases to estimate |
| `weights` | `c(ws=1, wz=1, wt=1, wc=1)` | Weights for spatial (`ws`), vertical (`wz`), temporal (`wt`), and cultural (`wc`) evidence. Since v0.14 they scale the feature dimensions and enter the likelihood, so they can be cross-validated with `optimize_weights()` |
| `em_iter` | 100 | Maximum EM iterations |
| `em_tol` | 1e-5 | Convergence tolerance (log-likelihood change) |
| `n_init` | 5 | Number of random initialisations (higher = more robust, slower) |
| `tafonomy` | NULL | Column name for taphonomic disturbance (0--1) |
| `context` | NULL | Column name for stratigraphic unit labels |
| `harris` | NULL | $n \times n$ penalty matrix (from `harris_from_contexts()` or `read_harris()`) |

### Statistical-model options

These options (all optional, sensible defaults) control the statistical core:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `class_model` | `"multinomial"` | How the material class is modelled. `"multinomial"` (recommended) treats it as a per-phase categorical distribution, so typology does not fragment stratigraphic units; `"gaussian"` is the legacy one-hot behaviour |
| `class_smoothing` | 1 | Dirichlet pseudo-count for the per-phase class probabilities (shrunk towards the global class frequencies) |
| `var_structure` | `"diagonal"` | `"diagonal"` (free per-dimension variances) or `"spherical"` (one shared variance per phase; the structure under which `weights` are identifiable) |
| `chrono_uncertainty` | FALSE | If TRUE, propagate each find's dating-interval width $(t_{max}-t_{min})^2/12$ into the temporal likelihood, so coarsely dated finds rely less on chronology |
| `strat_dynamic` | FALSE | If TRUE (with `context`), apply the stratigraphic constraint as a Neighborhood-EM field recomputed each E-step instead of a static penalty frozen at initialisation |
| `strat_beta` | 1 | Strength of the dynamic stratigraphic field |
| `noise` | FALSE | If TRUE, add a uniform background component (Fraley and Raftery, 1998); `$noise_prob` becomes a genuine outlier probability and extreme finds are kept out of the phase estimates |
| `noise_prior` | 0.05 | Initial mixing weight of the noise component |
| `chrono_precision` | FALSE | Add $1/t_{span}$ as a feature (precise dates weighted higher) |
| `taf_as_feature` | FALSE | Add `taf_score` as an explicit feature dimension |
| `residuality` | FALSE | Add a per-find date-vs-context-mean mismatch feature (needs `context`) |
| `class_scale` | FALSE | Scale one-hot class columns by $1/\sqrt{n_{classes}}$ (only relevant with `class_model = "gaussian"`) |
| `subclass` | NULL | Optional sub-class column used instead of `class` for the categorical/one-hot block |

`recommend_setup(data, context = "context", tafonomy = "taf_score")` inspects a dataset and suggests which of these to enable, flagging caveats such as unit-level coordinates or unit-tied chronology.

### Output: the `sef_fit` object

The returned object contains:

| Field | Type | Description |
|-------|------|-------------|
| `$phase` | integer vector | Dominant phase assignment per find |
| `$phase_prob` | matrix | $n \times k$ matrix of phase membership probabilities (conditional on not being noise when `noise = TRUE`) |
| `$cat_prob` | matrix | $k \times L$ per-phase class profile (multinomial model); see `phase_composition()` |
| `$noise_prob` | numeric vector | Per-find outlier posterior (when `noise = TRUE`) |
| `$entropy` | numeric vector | Shannon entropy per find |
| `$energy` | numeric vector | ESE per find |
| `$local_sei` | numeric vector | Local SEI per find |
| `$converged` | logical | Whether EM converged |
| `$n_init` | integer | Number of initialisations used |
| `$em_loglik` | numeric vector | Log-likelihood trace across iterations |
| `$model_stats` | list | Summary statistics (PDI, BIC, ICL, loglik, etc.) |

The per-phase class composition can be read directly:

```{r}
# Per-phase typological profile (multinomial model)
phase_composition(fit)
```

## Phase count selection: `compare_k()`

When $K$ is unknown, compare multiple values:

```{r}
ck <- compare_k(moderate, k_values = 2:5, seed = 42,
                tafonomy = "taf_score", context = "context")
print(ck)
```

The optimal $K$ minimises BIC while maximising PDI.

## Key diagnostics

```{r}
# Palimpsest Dissolution Index (global separability, 0--1)
pdi(fit)

# Compact summary
sef_summary(fit)

# Phase probability table
head(as_phase_table(fit))

# Full diagnostic table (input + diagnostics)
head(phase_diagnostic_table(fit))
```

# Intrusion Detection

`detect_intrusions()` returns, per find, an `intrusion_prob`, a `direction`
factor (`older_than_context` / `in_context` / `younger_than_context`), and a
signed `chrono_gap` (years). When the model was fitted with `noise = TRUE`,
`intrusion_prob` is the noise-component posterior — a genuine probability of
being an outlier; otherwise it is a heuristic composite of entropy, energy, and
inverse local SEI.

```{r}
fit_n <- fit_sef(moderate, k = 3, tafonomy = "taf_score",
                 context = "context", noise = TRUE, seed = 42)
di <- detect_intrusions(fit_n)
suspects <- di[di$intrusion_prob > 0.5, ]
cat("Suspected intrusions:", nrow(suspects), "\n")
head(suspects[order(suspects$intrusion_prob, decreasing = TRUE), ])
```

The `direction` distinguishes residual material (`older_than_context`) from
possible latent features (`younger_than_context`); it is informative only when
finds within a unit carry distinct chronologies (not under unit-tied dating).

Since v0.19, `detect_intrusions()` also returns an `intrusion_type` factor that
couples the intrusion magnitude with the direction: `residual` (flagged and
older-than-context), `latent_feature` (flagged and younger-than-context),
`outlier_in_context` (flagged but undirected), or `not_flagged`. `gg_outliers()`
colours finds by this type.

# Model Validation

When true phase labels are known:

```{r}
# Adjusted Rand Index (-1 to 1, 1 = perfect)
adjusted_rand_index(fit, moderate$true_phase)

# Confusion matrix (auto-reordered for best diagonal match)
confusion_matrix(fit, moderate$true_phase)
```

# Structured Exports

## Summary per stratigraphic unit

```{r}
us_summary_table(fit)
```

Columns: `context`, `n_finds`, `dominant_phase`, `purity`, `mean_entropy`, `mean_energy`, `mean_local_sei`, `n_intrusions`.

## Phase transition matrix

Shows how often finds from phase $i$ sit directly above finds from phase $j$:

```{r}
phase_transition_matrix(fit)
```

## Export all results to CSV

```{r eval=FALSE}
export_results(fit, dir = "results/", format = "csv", prefix = "site_X")
```

Creates 4 files:

- `site_X_phases.csv` --- phase assignments and diagnostics per find
- `site_X_intrusions.csv` --- intrusion probabilities
- `site_X_us_summary.csv` --- per-US aggregated statistics
- `site_X_model_summary.csv` --- model-level metrics (n, k, PDI, BIC, etc.)

# Harris Matrix Integration

## Auto-generate from depth ordering

```{r}
H <- harris_from_contexts(moderate, z_col = "z", context_col = "context",
                           penalty_scale = 0.5)
dim(H)
```

## Import from CSV

The CSV must have columns `from`, `to`, and optionally `weight`:

```
from,to,weight
SU_1,SU_2,1
SU_2,SU_3,1
SU_1,SU_3,0.5
```

```{r eval=FALSE}
H <- read_harris("harris_matrix.csv", contexts = data$context)
fit <- fit_sef(data, k = 3, harris = H, context = "context")
```

## Validate phases against stratigraphy

```{r}
validate_phases_harris(fit)
```

The `consistent` column flags whether the dominant phase ordering respects stratigraphic depth. `FALSE` indicates an inversion that may warrant investigation.

# Visualization

## ggplot2 plots (static)

All `gg_*` functions require **ggplot2** and return ggplot objects.

```{r fig.height=5}
library(ggplot2)

# Phase assignment map
gg_phasefield(fit)
```

```{r fig.height=5}
# Entropy map (uncertainty)
gg_entropy(fit)
```

```{r fig.height=5}
# Energy map (local disruption)
gg_energy(fit)
```

```{r fig.height=5}
# Intrusion detection map
gg_intrusions(fit, top_n = 5)
```

```{r fig.height=4}
# Phase count selection diagnostics
gg_compare_k(ck)
```

```{r fig.height=4}
# EM convergence trace
gg_convergence(fit)
```

```{r fig.height=5}
# Vertical phase profile (depth vs horizontal)
gg_phase_profile(fit)
```

```{r fig.height=5}
# Confusion matrix heatmap
gg_confusion(fit, moderate$true_phase)
```

### Model-output diagnostic plots (v0.18)

Four plots visualise the outputs of the refined statistical model:

```{r fig.height=4}
# Per-phase class composition (multinomial cat_prob), heatmap by default
gg_phase_composition(fit)
```

```{r eval=FALSE}
# Within-unit phase coherence; pass compare = a gaussian-class fit to contrast
gg_unit_coherence(fit)

# Non-spatial ranking of the intrusion probability (noise posterior if noise=TRUE)
gg_outliers(fit_n)

# Directional intrusions: residual vs latent (needs per-find dating)
gg_direction(fit_n)
```

### Type longevity

`type_longevity()` aggregates the posterior to estimate the temporal envelope
of each material class; `gg_longevity()` draws it as a Gantt-style chart:

```{r eval=FALSE}
tl <- type_longevity(fit)
gg_longevity(tl)
```

## Interactive plots (plotly)

```{r eval=FALSE}
library(plotly)

# Convert any gg_* plot to interactive
as_plotly(gg_phasefield(fit))
as_plotly(gg_entropy(fit))
as_plotly(gg_intrusions(fit))
```

Hover tooltips show: ID, context, phase probability, dating, class, entropy, energy, intrusion score.

## Base R plots

For environments without ggplot2:

```{r fig.height=5}
plot_phasefield(fit)
plot_entropy(fit)
plot_energy(fit)
plot_sei_profile(fit)
```

## GIS export

```{r eval=FALSE}
library(sf)

# Point layer with phase assignments
sf_pts <- as_sf_phase(fit, crs = 32632L, dims = "XYZ")
st_write(sf_pts, "results.gpkg", layer = "phases")

# High-SEI links as lines
sf_links <- as_sf_links(fit, quantile_threshold = 0.9, crs = 32632L)
st_write(sf_links, "results.gpkg", layer = "sei_links")
```

# Interpretive Report

Generate a plain-language report for inclusion in excavation documentation:

```{r}
report_sef(fit, lang = "en")
```

```{r}
report_sef(fit, lang = "it")
```

# Database Import

Import data from SQLite or PostgreSQL:

```{r eval=FALSE}
# SQLite
data <- read_db(
  con = DBI::dbConnect(RSQLite::SQLite(), "excavation.sqlite"),
  table = "finds"
)

# PostgreSQL (PyArchInit)
data <- read_db(
  con = DBI::dbConnect(RPostgres::Postgres(), dbname = "pyarchinit", host = "localhost"),
  table = "finds_table"
)
```

# Phase Reordering

Mixture models suffer from *label switching*: phase numbers are arbitrary across runs. Use `reorder_phases()` to ensure phase 1 = deepest (oldest) and phase K = shallowest (most recent):

```{r}
fit <- reorder_phases(fit)
table(fit$phase)
```

# Bootstrap Confidence Intervals

Assess uncertainty on key metrics via bootstrap resampling:

```{r}
bs <- bootstrap_sef(fit, n_boot = 50, verbose = FALSE)
print(bs)
```

Returns a data.frame with columns `metric`, `mean`, `se`, `ci_lower`, `ci_upper` for PDI, mean entropy, mean energy, loglik, and optionally ARI (if `true_labels` is provided).

Parameters:

| Parameter | Default | Description |
|-----------|---------|-------------|
| `n_boot` | 100 | Number of bootstrap replicates |
| `conf` | 0.95 | Confidence level for intervals |
| `true_labels` | NULL | Known true phases for ARI confidence interval |
| `verbose` | TRUE | Print progress messages |

# Cross-Validation

Evaluate model fit via k-fold cross-validation:

```{r}
cv <- cv_sef(moderate, k_values = 3, n_folds = 3, seed = 42)
print(cv)
```

Use with multiple K values to find the best:

```{r eval=FALSE}
cv <- cv_sef(data, k_values = 2:5, n_folds = 5, seed = 42)
print(cv)  # shows mean_test_loglik per k
```

# Data-Driven Weight Optimisation

Instead of manually tuning SEI weights, use `optimize_weights()` to find the best configuration via grid search:

```{r eval=FALSE}
opt <- optimize_weights(data, k = 3, n_folds = 3, seed = 42)
print(opt$best_weights)

# Re-fit with optimal weights
fit <- fit_sef(data, k = 3, weights = opt$best_weights)
```

# Sparse SEI for Large Datasets

For datasets with n > 1000, the full $n \times n$ SEI matrix may be too large. Use `sei_sparse()` which automatically applies a distance threshold:

```{r eval=FALSE}
# Automatic sparse mode
sei <- sei_sparse(data, max_dist = 5.0)

# Or use sei_matrix() directly with max_dist
sei <- sei_matrix(data, max_dist = 5.0)
```

Pairs beyond `max_dist` Euclidean distance receive SEI = 0, dramatically reducing memory and computation time.

# Complete Workflow Example

```{r eval=FALSE}
library(palimpsestr)

# 1. Load data
data <- read.csv("site_data.csv")

# 2. Select optimal K
ck <- compare_k(data, k_values = 2:6, context = "context",
                tafonomy = "taf_score", n_init = 3)
gg_compare_k(ck)
best_k <- ck$k[which.min(ck$bic)]

# 3. Fit with optimal K and multiple inits
fit <- fit_sef(data, k = best_k, context = "context",
               tafonomy = "taf_score", n_init = 5, seed = 42)

# 4. Inspect results
print(fit)
gg_phasefield(fit)
gg_entropy(fit)
gg_convergence(fit)
gg_phase_profile(fit)

# 5. Intrusion detection
gg_intrusions(fit, top_n = 10)

# 6. Harris validation
validate_phases_harris(fit)

# 7. Export
export_results(fit, dir = "results/")
report_sef(fit, lang = "en")

# 8. GIS
sf_pts <- as_sf_phase(fit, crs = 32632L)
sf::st_write(sf_pts, "results.gpkg", layer = "phases")
```

# Real-Data Example: Villa Romana

The `villa_romana` dataset is derived from the multi-period Roman villa of Poggio Gramignano (Lugnano in Teverina, Italy): 615 inventoried finds across 54 stratigraphic units and 17 material classes, spanning the Pre-Roman to Late Antique periods, with excavator-assigned taphonomic scores. Note that coordinates and dating are recorded at stratigraphic-unit level (see `recommend_setup()`), so the model resolves phases at the unit and macro-event level.

## Load and inspect

```{r}
data(villa_romana)
cat("Finds:", nrow(villa_romana), "| Contexts:", length(unique(villa_romana$context)),
    "| Classes:", length(unique(villa_romana$class)), "\n")
cat("Date range:", min(villa_romana$date_min), "to", max(villa_romana$date_max), "\n")
table(villa_romana$class)
```

## Fit with taphonomic weighting

```{r}
fit_vr <- fit_sef(villa_romana, k = 4,
                  tafonomy = "taf_score",
                  context  = "context",
                  n_init   = 5, seed = 42)
print(fit_vr)
```

## Diagnostics

```{r}
cat("PDI:", pdi(fit_vr), "\n")
sef_summary(fit_vr)
```

## Validation (on simulated data)

Validation with ARI and confusion matrix requires known true phases. We demonstrate with `demo_moderate`:

```{r}
data(demo_moderate)
fit_demo <- fit_sef(demo_moderate, k = 3, seed = 42)
adjusted_rand_index(fit_demo, demo_moderate$true_phase)
confusion_matrix(fit_demo, demo_moderate$true_phase)
```

## Intrusion detection

```{r}
di_vr <- detect_intrusions(fit_vr)
suspects_vr <- di_vr[di_vr$intrusion_prob > 0.5, ]
cat("Suspected intrusions:", nrow(suspects_vr), "out of", nrow(villa_romana), "\n")
```

## Visualisation

```{r fig.height=5}
gg_phasefield(fit_vr)
```

```{r fig.height=5}
gg_entropy(fit_vr)
```

```{r fig.height=5}
gg_energy(fit_vr)
```

```{r fig.height=5}
gg_intrusions(fit_vr, top_n = 10)
```

```{r fig.height=5}
gg_phase_profile(fit_vr)
```

## Per-US summary

```{r}
us_summary_table(fit_vr)
```

## Interpretive report

```{r}
report_sef(fit_vr, lang = "en")
```

# Function Reference

## Core

| Function | Description |
|----------|-------------|
| `fit_sef()` | Fit the SEF model |
| `compare_k()` | Compare multiple K values |
| `archaeo_sim()` | Simulate archaeological data |
| `reorder_phases()` | Relabel phases by mean depth (deepest = phase 1) |

## Diagnostics

| Function | Description |
|----------|-------------|
| `pdi()` | Palimpsest Dissolution Index |
| `sef_summary()` | Compact model summary |
| `as_phase_table()` | Phase probability table |
| `predict_phase()` | Alias for `as_phase_table()` |
| `phase_diagnostic_table()` | Full diagnostic table |
| `detect_intrusions()` | Intrusion detection |

## Validation

| Function | Description |
|----------|-------------|
| `adjusted_rand_index()` | Adjusted Rand Index |
| `confusion_matrix()` | Confusion matrix (auto-reordered) |

## Export

| Function | Description |
|----------|-------------|
| `us_summary_table()` | Per-US summary statistics |
| `phase_transition_matrix()` | Vertical transition counts |
| `export_results()` | Export all results to CSV |

## Harris Matrix

| Function | Description |
|----------|-------------|
| `harris_from_contexts()` | Auto-generate penalty from depths |
| `read_harris()` | Import Harris Matrix from CSV |
| `validate_phases_harris()` | Check phase--stratigraphy consistency |

## Robustness & Sensitivity

| Function | Description |
|----------|-------------|
| `bootstrap_sef()` | Bootstrap confidence intervals on PDI, entropy, ARI |
| `cv_sef()` | K-fold cross-validation for model selection |
| `optimize_weights()` | Data-driven SEI weight optimisation via grid search |

## SEI / ESE

| Function | Description |
|----------|-------------|
| `sei_matrix()` | Stratigraphic Entanglement Index matrix |
| `sei_sparse()` | Sparse SEI for large datasets (n > 1000) |
| `local_sei()` | Row sums of SEI matrix |
| `ese()` | Excavation Stratigraphic Energy |

## Plots (ggplot2)

| Function | Description |
|----------|-------------|
| `gg_phasefield()` | Phase assignment map |
| `gg_entropy()` | Entropy map |
| `gg_energy()` | ESE energy map |
| `gg_intrusions()` | Intrusion detection map |
| `gg_compare_k()` | K selection diagnostics |
| `gg_convergence()` | EM convergence trace |
| `gg_phase_profile()` | Vertical depth profile |
| `gg_confusion()` | Confusion matrix heatmap |
| `gg_cv()` | Cross-validation diagnostic plot |
| `gg_bootstrap()` | Bootstrap confidence interval plot |
| `gg_weights()` | Weight sensitivity heatmap |
| `gg_map()` | Overlay on excavation plan |
| `as_plotly()` | Convert any gg_* to interactive |

## Plots (base R)

| Function | Description |
|----------|-------------|
| `plot_phasefield()` | Phase map |
| `plot_entropy()` | Entropy map |
| `plot_energy()` | Energy map |
| `plot_sei_profile()` | SEI profile |

## GIS

| Function | Description |
|----------|-------------|
| `as_sf_phase()` | Export as sf point layer |
| `as_sf_links()` | Export SEI links as sf lines |
| `load_geometries()` | Load excavation plan geometries |

## Other

| Function | Description |
|----------|-------------|
| `report_sef()` | Interpretive report (en/it) |
| `read_db()` | Import from SQLite/PostgreSQL |
